---
title: "miRPM Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{miRPM Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

`miRPM` provides a transparent workflow for exploratory analysis of
miRNA-seq count matrices using reads per million (RPM) normalization.

The workflow includes:

1. RPM normalization.
2. Abundance and prevalence filtering.
3. Nonparametric group comparisons.
4. Static and interactive expression plots.

RPM is an abundance-scaling method. It does not replace count-based
differential-expression models such as DESeq2, edgeR, or limma-voom
when modelling complex experimental designs, composition bias, or
count-dependent variance is required.

## Load the package and example data

```{r load-package}
library(miRPM)

data("miRPM_example", package = "miRPM")

counts <- miRPM_example$counts
metadata <- miRPM_example$metadata
```

The example dataset is synthetic and is intended only for
documentation and testing.

```{r inspect-data}
dim(counts)
head(counts)
metadata
```

The count matrix must contain miRNAs in rows and samples in columns.

```{r validate-alignment}
stopifnot(
  identical(colnames(counts), metadata$Sample),
  all(counts >= 0),
  all(is.finite(counts))
)
```

## Normalize counts to RPM

By default, `normalize_rpm()` calculates each library size from the
complete column sum of the input count matrix.

```{r normalize}
rpm <- normalize_rpm(counts)

dim(rpm)
round(rpm[, 1:4], 2)
```

The returned matrix contains normalization metadata.

```{r normalization-metadata}
attr(rpm, "miRPM_normalization")
```

Externally calculated library sizes can also be supplied as a named
vector.

```{r external-library-sizes}
library_sizes <- setNames(
  c(
    2.1e6, 2.3e6, 2.0e6, 2.2e6,
    1.9e6, 2.4e6, 2.1e6, 2.0e6
  ),
  colnames(counts)
)

rpm_external <- normalize_rpm(
  count_matrix = counts,
  library_sizes = library_sizes
)

dim(rpm_external)
```

## Filter low-information miRNAs

Filtering can combine:

- Minimum RPM abundance.
- Minimum raw count.
- Minimum prevalence.
- Minimum number of qualifying samples.
- Group-specific prevalence requirements.

The following example retains miRNAs satisfying the criteria in at
least one experimental group.

```{r filter}
filtered <- filter_mirnas(
  rpm_matrix = rpm,
  metadata = metadata,
  raw_counts = counts,
  min_rpm = 5,
  min_count = 10,
  min_prevalence = 0.5,
  min_samples = 2,
  group_column = "Condition",
  prevalence_scope = "any_group",
  return_diagnostics = TRUE
)

filtered$summary
```

The filtered RPM matrix is available in
`filtered$filtered_matrix`.

```{r inspect-filtered}
rownames(filtered$filtered_matrix)
dim(filtered$filtered_matrix)
```

Detailed filtering information is available for each miRNA.

```{r filtering-diagnostics}
filtered$diagnostics
```

Filtering thresholds should be chosen according to sequencing depth,
sample size, biological context, and the intended downstream analysis.
They should not be interpreted as universal detection thresholds.

## Perform nonparametric group comparisons

`perform_statistical_tests()` analyzes the filtered RPM matrix using
nonparametric methods.

For two independent groups, the function performs a Wilcoxon rank-sum
test. For more than two groups, it performs a Kruskal-Wallis test and
can calculate Dunn post-hoc comparisons.

```{r statistical-analysis}
de_result <- perform_statistical_tests(
  miRNA_ftd = filtered$filtered_matrix,
  metadata = metadata,
  condition = "Condition",
  group_order = c("Control", "Disease"),
  output_file = NULL,
  assign_results = FALSE,
  write_excel = FALSE
)
```

The complete result table includes raw and adjusted p-values together
with descriptive and effect-size information.

```{r statistical-results}
de_result$results
```

Results passing the selected significance threshold are stored
separately.

```{r significant-results}
de_result$significant_results
```

Adjusted p-values control the false discovery rate across the tested
miRNAs. Statistical significance should be interpreted together with
effect size, abundance, prevalence, consistency, and study design.

## Create a joint expression plot

`miRNA_expression_plot()` returns both a static `ggplot2` object and an
interactive `plotly` object.

```{r joint-plot}
joint_plot <- miRNA_expression_plot(
  mirnas = rownames(filtered$filtered_matrix),
  rpm_matrix = filtered$filtered_matrix,
  metadata = metadata,
  condition_column = "Condition",
  sample_column = "Sample",
  groups = c("Control", "Disease"),
  save_html = FALSE
)

joint_plot$ggplot
```

The interactive object is available as:

```{r interactive-object}
class(joint_plot$interactive)
```

It can be displayed interactively in an RStudio viewer, R Markdown
document, or compatible HTML environment.

## Create individual miRNA plots

`individual_miRNA_plot()` creates one static plot for each selected
miRNA.

```{r individual-plots}
individual_plots <- individual_miRNA_plot(
  filtered_results = de_result$results,
  rpm_matrix = filtered$filtered_matrix,
  metadata = metadata,
  condition_column = "Condition",
  sample_column = "Sample",
  groups_to_include = c("Control", "Disease"),
  adjusted_pvalue_column = "FDR",
  save_png = FALSE
)

length(individual_plots$plots)
```

Display the first plot:

```{r first-individual-plot}
individual_plots$plots[[1]]
```

The returned object also reports selected miRNAs, missing miRNAs,
adjusted p-values, data used for plotting, and any files written to
disk.

```{r individual-output}
names(individual_plots)
individual_plots$adjusted_pvalues
```

## Reproducible workflow summary

A typical `miRPM` analysis follows this order:

```{r workflow-summary, eval=FALSE}
rpm <- normalize_rpm(counts)

filtered <- filter_mirnas(
  rpm_matrix = rpm,
  metadata = metadata,
  raw_counts = counts,
  group_column = "Condition",
  prevalence_scope = "any_group",
  return_diagnostics = TRUE
)

results <- perform_statistical_tests(
  miRNA_ftd = filtered$filtered_matrix,
  metadata = metadata,
  condition = "Condition",
  assign_results = FALSE,
  write_excel = FALSE
)
```

Normalization should be performed before filtering so that library
sizes are calculated from the complete count matrix unless valid
external library sizes are supplied.
